---
title: "Example R script"
author: "CRG"
date: "May  2019"
output: html_document
---

The following R script is illustrative of the workflow I follow in R as a Monitoring, Evaluation, and Learning (MEL) Advisor at IREX. This example rmarkdown document pulls from an analysis I did for a youth-led labor market assessment in Southern Africa. 

```{r}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  results = 'asis'
)
```

```{r}
# load packages
library(pacman)
p_load(tidyverse, sjmisc, janitor, knitr, ggthemes, corrplot, corrr, summarytools, corrplot, ggpubr, ggalt, RCurl, viridis)
```
# load data  
```{r load data}
youth <- read.csv(text = getURL("https://raw.githubusercontent.com/cguedenet/sample/master/youth2.csv"))
```

#examine data
```{r examine data, eval = FALSE}
empty <- names(empty_cols(youth))
youth %>% remove_cols(empty) %>% data.frame() %>% glimpse()
```


# data cleaning using janitor package
```{r data cleaning, eval = TRUE, echo = FALSE}
# change all empty cells to na values using sjmisc package
youth <- youth %>% set_na(na = "", drop.levels = TRUE)

## Any empty columns or rows?
empty_cols(youth)
empty_rows(youth)

##remove all empty rows
youth <- youth %>% remove_empty("rows")

##remove all numeric responses for "region" variable.These were submissions from an early pilot version of the survey
youth <- youth %>% filter(!region %in% c(5,6,8)) %>%
  droplevels(.)

##frequency table 
table(youth$region)
```

# recode data  
```{r recoding}
# group response options: For tenure question (#38), recode all responses that are 6 or greater as "6 or more"
youth <- youth %>% mutate(number_of_years = as.factor(case_when(
  .$number_of_years == "6" ~ "6 or more",
  .$number_of_years == "7" ~ "6 or more",
  .$number_of_years == "8" ~ "6 or more",
  .$number_of_years == "9" ~ "6 or more",
  .$number_of_years == "10" ~ "6 or more",
  .$number_of_years == "11" ~ "6 or more",
  TRUE ~ as.character(.$number_of_years)
)))

# create new variable: employ.status
youth <- youth %>% mutate(employ.status = ifelse(youth$status == "Not_currently_working", "unemployed", "employed"))

# create new variable - Number of years since finishing studies
youth <- youth %>% mutate(yearsFinished = 2018 - year_studies)
```

#rename variables
```{r rename}
#rename variables with employment obstacles
youth <- rename(youth, Obs_No_education = obstacles_1,
                 Obs_Unsuitable_general_ed=obstacles_2,
                 Obs_Unsuitable_vocational_ed=obstacles_3,
                 Obs_No_suitable_training=obstacles_4,
                 Obs_Requirements=obstacles_5,
                 Obs_training_received=obstacles_6,
                 Obs_No_work_experience=obstacles_7,
                 Obs_Not_enough_jobs=obstacles_8,
                 Obs_Considered_too_young=obstacles_9,
                 Obs_Being_male.female=obstacles_10,
                 Obs_Discriminatory_prej=obstacles_11,
                 Obs_Low_wages=obstacles_12,
                 Obs_Poor_working_conditions=obstacles_13,
                 Obs_Disability=obstacles_14,
                 Obs_Health_issues=obstacles_15,
                 Obs_Other2=obstacles_16)
```

# check for missing values  
```{r missing values}
# create new column with totals for number of missing responses; filter to include only those with more 100 NAs (i.e. about 40% of all questions left blank)
youth <- youth %>% row_count(count = NA, var = "rowcount", append = TRUE)

# quick check on number and percentage of missing responses per column
youth %>%
  tabyl(rowcount) %>%
  adorn_pct_formatting(digits = 2) %>%
  arrange(desc(rowcount)) %>%
  kable()

## why do some surveys have so many missing values?
descr(youth$rowcount) # 25% of surveys have more than 65 missing values

# employed and unemployed youth got different questions
```

#Check for outliers for question about the number of jobs youth had applied for in the last 12 months
```{r outliers}

kable(frq(youth$jobs))

#First, remove extreme outliers that overly influence mean and SD
youth.jobs <- youth %>%
  filter(jobs < 100)

# create boxplot
boxplot(youth.jobs$jobs)

#isolate outliers (defining this as all values at least 3 standard deviations from mean)
outliers <- boxplot(youth.jobs$jobs, range = 3, plot = FALSE)$out

#remove outliers
youth.jobs <- youth.jobs %>% filter(!jobs %in% outliers)

#summary table
youth.jobs %>% frq(jobs) %>% kable()
```

#Tabulations
```{r tabulations, results = 'asis'}
# cross-tabulation of youth education and employment status
youth %>%
  flat_table(employ.status, education) %>%
  data.frame() %>%
  spread(employ.status, Freq) %>%
  kable()

# cross-tabulation of region and discriminatory prejudices
youth %>%
  flat_table(region, Obs_Discriminatory_prej) %>%
  data.frame() %>%
  kable()
```

#Analysis
###Comparison of obstacles cited by employed and unemployed  
```{r analysis}
youth %>%
  select("employ.status",85:100) %>%
  gather("obstacles", "n",2:17) %>%
  filter(n==1) %>%
  tabyl(obstacles, employ.status) %>%
  mutate(difference = abs(employed - unemployed)) %>%
  adorn_totals("col") %>%
  arrange(desc(Total)) %>%
  kable()
```

##To what extent do you believe these identities affect people's ability to get a job?
```{r analysis2}
identities <- data.frame(sapply(youth[, 103:107], flat_table, margin = "row")) %>%
  filter(Gender != 0) %>%
  mutate(name = c("A great deal", "Never", "Occasionally")) %>%
  select(c(6, 1:5)) %>%
  .[c(2, 3, 1), ]
 
kable(identities)
```

#visualize contingency table using Balloon plot from ggpubr package
```{r }
#create matrix-like data frame with row names
row.names(identities) <- identities$name 
identities <- identities[,-1]

# balloon plot
ggballoonplot(identities, fill = "value", color = "lightgray",
              show.label = TRUE, font.label = c(12,"plain","white"),
              size = 14) +
  scale_fill_viridis(direction = -1, option = "inferno") +
  theme_minimal() +
  theme(axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.text = element_text(size = 10)) +
  theme(panel.grid.major = element_line(color = "#d8e6e6"),
        panel.grid.minor = element_line("#d8e6e6")) +
  labs(title = "frequency table", x = "identities", y = "response options")
```

## association between prejudice as an obstacle and region  
```{r}
## gather obstacles in long format
obstaclesAll <- youth %>%
  select(85:100) %>%
  gather("obstacles", "sum") %>%
  group_by(obstacles) %>%
  summarise(total = sum(sum, na.rm = TRUE)) %>%
  arrange(desc(total))

#youth and prejudice obstacle crosstabs. There may be an association between youth from region 3 and the belief that regional origin is an obstacle to getting a job
youth %>% flat_table(reg_origina,region, margin = "col") %>% data.frame() %>%
  spread(region, Freq) %>%
  kable()

##Pearson's Chi-squared Test with stats package
#association between a youth's home region and the belief that region as affecting a person's ability to get a job
corrTable <- chisq.test(youth$region, youth$reg_origina)

corrTable

#look at residuals
corrplot(corrTable$residuals, is.cor = FALSE, tl.col = "black", cl.ratio = 0.2, cl.align.text = "l")
```



##minimum income
```{r geomerror, fig.height=4, fig.width=7}
# Is there a minimum income below which you would not accepted a job?

# comparison with education
incomel <- youth %>%
  group_by(education) %>%
  summarise(mn = mean(min_incomelevel, na.rm = TRUE), med = median(min_incomelevel, na.rm = TRUE), dist = n_distinct(min_incomelevel), min = min(min_incomelevel, na.rm = TRUE), max = max(min_incomelevel, na.rm = TRUE), sd = sd(min_incomelevel, na.rm = TRUE), IQR = IQR(min_incomelevel, na.rm = TRUE)) %>%
  data.frame() %>%
   na.omit(.)

## get rid of underscores (converts to character format)
incomel$education <- gsub("_", " ", incomel$education)

## create nice looking plot
labels <- incomel$education
averages <- incomel$mn
ranges <- incomel$IQR

# geom_point plot with error bars
plot1 <- ggplot(incomel, aes(averages, labels)) + geom_point(aes(color = "red", size = .8)) + scale_x_continuous(labels = scales::comma) +
  geom_errorbarh(aes(xmin = averages - ranges, xmax = averages + ranges, height = .2, width = .3), color = "black") +
  labs(title = "Minimum income levels youth would accept", subtitle = "Range and average values", x = "Minimum Acceptable Income", y = "Education Level") + theme(
    axis.text = element_text(size = 11),
    axis.text.x = element_text(size = 11),
    plot.title = element_text(size = 11, face = "bold", hjust = 2),
    plot.subtitle = element_text(hjust = -2))

plot1 <-  plot1 +  geom_hline(yintercept = 0, size =1, colour = "#9ba7a7") +
  theme_minimal() +
  theme(axis.ticks.x=element_blank(),
        panel.grid.major = element_line(color = "#d8e6e6" ),
        panel.grid.minor = element_line("#d8e6e6"),
        axis.title = element_text(face = "bold"),
        legend.position = "none")

print(plot1)
```

#visualizations

```

##lollipop charts
```{r, fig.width = 10}
library(ggalt)
lolli <- youth %>% select(ID,status, region) %>%
  na.omit(.) %>% data.frame() %>%
  add_count(region, name = "regFreq") %>%
  add_count(status,region, name = "status.regFreq") %>%
  mutate(region2 = paste(region," ","(n=", regFreq,")"))
  
#take out underscores periods in status variable
lolli$status <- gsub("[_.]", " ", lolli$status)

#reorder status variable from least to greatest n
lolli$status <- fct_reorder2(lolli$status,lolli$regFreq,lolli$status.regFreq)

# lollipo chart
lolliPlot <- ggplot(lolli, aes(status, status.regFreq)) +
  geom_lollipop(point.colour="#098b8e", point.size=5) +
  coord_flip() +
  facet_wrap(~region2)

#add theme elements
lolliPlot + scale_y_continuous(expand=c(0,0),limits=c(0, 55)) +
  coord_flip() +
  labs(x=NULL, y=NULL) +
  theme_minimal() +
  theme(panel.grid.major.y=element_blank()) +
  theme(panel.grid.minor=element_blank()) +
  theme(axis.line.y=element_line(color="#9ba7a7", size=0.15))+
  theme(axis.text.y=element_text(margin=margin(r=5,l=0),size=10,face="bold"))+
  theme(axis.text.x=element_text(size=11))+
  theme(plot.margin=unit(c(30,30,30,30),"pt"))
```
